### 2025-02-10 paper/preprint/paper.tex
progress

\subsection{Encoding Actionability
Constraints}\label{encoding-actionability-constraints}

@@

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. This is best
explained through a simple example involving a linear classifier:
\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
\(y^+ \sim p(y)\) we have:

\begin{equation}\phantomsection\label{eq-grad}{
\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
}\end{equation}

In other words, contrastive divergence will push down (up) on the
coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
(higher) for counterfactuals than samples. This has the intended effect
of aligning counterfactual explanations with the data, as discussed
above.

Crucially though, we want to avoid penalizing implausibility that arises
due to mutability constraints. Suppose, for example, that feature \(d\)
represents \emph{age} and our model predicts the likelihood of credit
default.

@@

\begin{itemize}
\tightlist
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\end{itemize}

@@

@@

\begin{itemize}
\tightlist
\item
  Generator Parameters:

  \begin{itemize}
  \tightlist
  \item
    \(\lambda_{\text{cost}}\): \texttt{0.0,\ 0.001,\ 0.1}
  \item
    \(\lambda_{\text{div}}\):
    \texttt{0.01,\ 0.05,\ 0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 15.0}
  \item
    Learning Rate: \texttt{1.0}
  \item
    Maximum Iterations: \texttt{20,\ 50,\ 100}
  \item
    Optimizerimizer: \texttt{sgd}
  \end{itemize}
\item
  Generator: \texttt{ecco,\ generic,\ omni,\ revise}
\item
  Training Parameters:

  \begin{itemize}
  \tightlist
  \item
    Objective: \texttt{full,\ vanilla}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Counterfactual Parameters:

  \begin{itemize}
  \tightlist
  \item
    Convergence: \texttt{max\_iter}
  \item
    Maximum Iterations: \texttt{100}
  \item
    No.~Individuals: \texttt{100}
  \item
    No.~Runs: \texttt{5}
  \end{itemize}
\item
  Generator Parameters:

  \begin{itemize}
  \tightlist
  \item
    \(\lambda_{\text{cost}}\): \texttt{0.0}
  \item
    \(\lambda_{\text{div}}\):
    \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 20.0}
  \item
    Learning Rate: \texttt{1.0}
  \item
    Maximum Iterations: \texttt{50}
  \item
    Optimizerimizer: \texttt{sgd}
  \end{itemize}
\end{itemize}

@@

\begin{longtable}{ccccc}

@@

\tabularnewline

  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  full & 0.01 & \textit{ECCo} & $-9.91 \cdot 10^{11}$ & $2.25 \cdot 10^{12}$ \\
  full & 0.01 & \textit{Generic} & $-5.71 \cdot 10^{17}$ & $1.3 \cdot 10^{18}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.116}} \\
  full & 0.01 & \textit{REVISE} & -15.6 & 13.2 \\
  vanilla & 0.01 & \textit{ECCo} & -4.28 & 3.52 \\
  vanilla & 0.01 & \textit{Generic} & -4.45 & 3.47 \\
  vanilla & 0.01 & \textit{Omniscient} & -5.12 & 4.46 \\
  vanilla & 0.01 & \textit{REVISE} & -4.91 & 4.24 \\
  full & 0.05 & \textit{ECCo} & $-5.63 \cdot 10^{5}$ & $1.28 \cdot 10^{6}$ \\
  full & 0.05 & \textit{Generic} & $-8.35 \cdot 10^{17}$ & $1.9 \cdot 10^{18}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.114}} \\
  full & 0.05 & \textit{REVISE} & -15 & 12.6 \\
  vanilla & 0.05 & \textit{ECCo} & -4.4 & 3.66 \\
  vanilla & 0.05 & \textit{Generic} & -4.38 & 3.48 \\
  vanilla & 0.05 & \textit{Omniscient} & -5.25 & 4.62 \\
  vanilla & 0.05 & \textit{REVISE} & -4.94 & 4.22 \\
  full & 0.1 & \textit{ECCo} & $-6.74 \cdot 10^{5}$ & $1.53 \cdot 10^{6}$ \\
  full & 0.1 & \textit{Generic} & $-1.72 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.56}} & \color{blue}{\textbf{0.124}} \\
  full & 0.1 & \textit{REVISE} & -15.6 & 13.2 \\
  vanilla & 0.1 & \textit{ECCo} & -4.28 & 3.52 \\
  vanilla & 0.1 & \textit{Generic} & -4.45 & 3.48 \\
  vanilla & 0.1 & \textit{Omniscient} & -5.12 & 4.46 \\
  vanilla & 0.1 & \textit{REVISE} & -4.91 & 4.25 \\
  full & 0.5 & \textit{ECCo} & -11.8 & 9.83 \\
  full & 0.5 & \textit{Generic} & $-1.06 \cdot 10^{18}$ & $2.42 \cdot 10^{18}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.123}} \\
  full & 0.5 & \textit{REVISE} & -15 & 12.6 \\
  vanilla & 0.5 & \textit{ECCo} & -4.4 & 3.65 \\
  vanilla & 0.5 & \textit{Generic} & -4.38 & 3.48 \\
  vanilla & 0.5 & \textit{Omniscient} & -5.25 & 4.61 \\
  vanilla & 0.5 & \textit{REVISE} & -4.95 & 4.22 \\
  full & 1 & \textit{ECCo} & -11.5 & 11.1 \\
  full & 1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.88 \cdot 10^{11}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.59}} & \color{blue}{\textbf{0.117}} \\
  full & 1 & \textit{REVISE} & -15.7 & 13.3 \\
  vanilla & 1 & \textit{ECCo} & -4.28 & 3.51 \\
  vanilla & 1 & \textit{Generic} & -4.44 & 3.47 \\
  vanilla & 1 & \textit{Omniscient} & -5.11 & 4.46 \\
  vanilla & 1 & \textit{REVISE} & -4.91 & 4.25 \\
  full & 5 & \textit{ECCo} & -3.99 & 3.12 \\
  full & 5 & \textit{Generic} & $-4.88 \cdot 10^{17}$ & $1.11 \cdot 10^{18}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.117}} \\
  full & 5 & \textit{REVISE} & -14.6 & 12.1 \\
  vanilla & 5 & \textit{ECCo} & -4.4 & 3.65 \\
  vanilla & 5 & \textit{Generic} & -4.38 & 3.48 \\
  vanilla & 5 & \textit{Omniscient} & -5.25 & 4.61 \\
  vanilla & 5 & \textit{REVISE} & -4.95 & 4.22 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.31}} & \color{blue}{\textbf{0.735}} \\
  full & 10 & \textit{Generic} & $-1.7 \cdot 10^{11}$ & $3.86 \cdot 10^{11}$ \\
  full & 10 & \textit{Omniscient} & -2.53 & 0.117 \\
  full & 10 & \textit{REVISE} & -15.5 & 13 \\
  vanilla & 10 & \textit{ECCo} & -4.28 & 3.51 \\
  vanilla & 10 & \textit{Generic} & -4.44 & 3.47 \\
  vanilla & 10 & \textit{Omniscient} & -5.12 & 4.46 \\
  vanilla & 10 & \textit{REVISE} & -4.91 & 4.24 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.01}} & \color{blue}{\textbf{0.488}} \\
  full & 15 & \textit{Generic} & $-4.91 \cdot 10^{17}$ & $1.12 \cdot 10^{18}$ \\
  full & 15 & \textit{Omniscient} & -2.53 & 0.116 \\
  full & 15 & \textit{REVISE} & -14.4 & 11.7 \\
  vanilla & 15 & \textit{ECCo} & -4.4 & 3.65 \\
  vanilla & 15 & \textit{Generic} & -4.38 & 3.48 \\
  vanilla & 15 & \textit{Omniscient} & -5.25 & 4.6 \\
  vanilla & 15 & \textit{REVISE} & -4.95 & 4.23 \\\bottomrule

\end{longtable}

\begin{longtable}{ccccc}

@@

\tabularnewline

  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  full & 0 & \textit{ECCo} & $-5.32 \cdot 10^{3}$ & $1.21 \cdot 10^{4}$ \\
  full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.64}} & \color{blue}{\textbf{0.125}} \\
  full & 0 & \textit{REVISE} & -15.4 & 12.9 \\
  vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0 & \textit{Omniscient} & -5.18 & 4.54 \\
  vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
  full & 0.001 & \textit{ECCo} & -362 & 811 \\
  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.03 \cdot 10^{17}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
  full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
  vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0.001 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0.001 & \textit{Omniscient} & -5.18 & 4.53 \\
  vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
  full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
  full & 0.1 & \textit{Generic} & $-4.49 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\
  full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
  vanilla & 0.1 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0.1 & \textit{Omniscient} & -5.18 & 4.54 \\
  vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule

\end{longtable}

@@

\tabularnewline

  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  full & 0.01 & \textit{ECCo} & $-2.8 \cdot 10^{22}$ & $6.39 \cdot 10^{22}$ \\
  full & 0.01 & \textit{Generic} & $-4.89 \cdot 10^{30}$ & $1.11 \cdot 10^{31}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.74}} & \color{blue}{\textbf{5.08}} \\
  full & 0.01 & \textit{REVISE} & -572 & $1.25 \cdot 10^{3}$ \\
  vanilla & 0.01 & \textit{ECCo} & -15.5 & 17.3 \\
  vanilla & 0.01 & \textit{Generic} & -10.9 & 11.9 \\
  vanilla & 0.01 & \textit{Omniscient} & -12.7 & 14.4 \\
  vanilla & 0.01 & \textit{REVISE} & -11.2 & 13 \\
  full & 0.05 & \textit{ECCo} & $-1.55 \cdot 10^{16}$ & $3.52 \cdot 10^{16}$ \\
  full & 0.05 & \textit{Generic} & $-2.22 \cdot 10^{20}$ & $5 \cdot 10^{20}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{4.48}} \\
  full & 0.05 & \textit{REVISE} & $-1.04 \cdot 10^{3}$ & $2.3 \cdot 10^{3}$ \\
  vanilla & 0.05 & \textit{ECCo} & -15.5 & 17.2 \\
  vanilla & 0.05 & \textit{Generic} & -11.7 & 12.8 \\
  vanilla & 0.05 & \textit{Omniscient} & -12.4 & 14.1 \\
  vanilla & 0.05 & \textit{REVISE} & -11.3 & 13.1 \\
  full & 0.1 & \textit{ECCo} & $-3.41 \cdot 10^{3}$ & $7.73 \cdot 10^{3}$ \\
  full & 0.1 & \textit{Generic} & $-5.22 \cdot 10^{30}$ & $1.19 \cdot 10^{31}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.78}} & \color{blue}{\textbf{5.12}} \\
  full & 0.1 & \textit{REVISE} & -288 & 594 \\
  vanilla & 0.1 & \textit{ECCo} & -15.5 & 17.2 \\
  vanilla & 0.1 & \textit{Generic} & -10.9 & 11.9 \\
  vanilla & 0.1 & \textit{Omniscient} & -12.7 & 14.4 \\
  vanilla & 0.1 & \textit{REVISE} & -11.3 & 13.1 \\
  full & 0.5 & \textit{ECCo} & -7.09 & 7.51 \\
  full & 0.5 & \textit{Generic} & $-1.11 \cdot 10^{31}$ & $2.53 \cdot 10^{31}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.58}} & \color{blue}{\textbf{4.83}} \\
  full & 0.5 & \textit{REVISE} & $-1.19 \cdot 10^{3}$ & $2.64 \cdot 10^{3}$ \\
  vanilla & 0.5 & \textit{ECCo} & -15.5 & 17.2 \\
  vanilla & 0.5 & \textit{Generic} & -11.7 & 12.8 \\
  vanilla & 0.5 & \textit{Omniscient} & -12.4 & 14.1 \\
  vanilla & 0.5 & \textit{REVISE} & -11.3 & 13.1 \\
  full & 1 & \textit{ECCo} & -6.06 & 6.33 \\
  full & 1 & \textit{Generic} & $-1.58 \cdot 10^{33}$ & $3.59 \cdot 10^{33}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.66}} & \color{blue}{\textbf{4.89}} \\
  full & 1 & \textit{REVISE} & $-1.16 \cdot 10^{3}$ & $2.59 \cdot 10^{3}$ \\
  vanilla & 1 & \textit{ECCo} & -15.5 & 17.3 \\
  vanilla & 1 & \textit{Generic} & -10.9 & 11.9 \\
  vanilla & 1 & \textit{Omniscient} & -12.7 & 14.4 \\
  vanilla & 1 & \textit{REVISE} & -11.3 & 13.1 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.57}} & \color{blue}{\textbf{2.07}} \\
  full & 5 & \textit{Generic} & $-1.17 \cdot 10^{28}$ & $2.66 \cdot 10^{28}$ \\
  full & 5 & \textit{Omniscient} & -4.29 & 4.31 \\
  full & 5 & \textit{REVISE} & -530 & $1.16 \cdot 10^{3}$ \\
  vanilla & 5 & \textit{ECCo} & -15.5 & 17.2 \\
  vanilla & 5 & \textit{Generic} & -11.7 & 12.7 \\
  vanilla & 5 & \textit{Omniscient} & -12.4 & 14.1 \\
  vanilla & 5 & \textit{REVISE} & -11.3 & 13.1 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.76}} & \color{blue}{\textbf{0.974}} \\
  full & 10 & \textit{Generic} & $-1.54 \cdot 10^{33}$ & $3.51 \cdot 10^{33}$ \\
  full & 10 & \textit{Omniscient} & -4.44 & 4.56 \\
  full & 10 & \textit{REVISE} & $-1.52 \cdot 10^{3}$ & $3.4 \cdot 10^{3}$ \\
  vanilla & 10 & \textit{ECCo} & -15.5 & 17.3 \\
  vanilla & 10 & \textit{Generic} & -10.9 & 11.9 \\
  vanilla & 10 & \textit{Omniscient} & -12.7 & 14.4 \\
  vanilla & 10 & \textit{REVISE} & -11.3 & 13.1 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.37}} & \color{blue}{\textbf{0.365}} \\
  full & 15 & \textit{Generic} & $-5.32 \cdot 10^{28}$ & $1.21 \cdot 10^{29}$ \\
  full & 15 & \textit{Omniscient} & -4.34 & 4.38 \\
  full & 15 & \textit{REVISE} & -473 & $1.03 \cdot 10^{3}$ \\
  vanilla & 15 & \textit{ECCo} & -15.5 & 17.2 \\
  vanilla & 15 & \textit{Generic} & -11.7 & 12.8 \\
  vanilla & 15 & \textit{Omniscient} & -12.4 & 14.1 \\
  vanilla & 15 & \textit{REVISE} & -11.3 & 13.1 \\\bottomrule

\end{longtable}

@@

\tabularnewline

  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.26}} & \color{blue}{\textbf{0.423}} \\
  full & 0.01 & \textit{Generic} & -1.49 & 0.71 \\
  full & 0.01 & \textit{Omniscient} & -5.21 & 5.25 \\
  full & 0.01 & \textit{REVISE} & $-2.71 \cdot 10^{26}$ & $6.37 \cdot 10^{26}$ \\
  vanilla & 0.01 & \textit{ECCo} & -9.33 & 7.34 \\
  vanilla & 0.01 & \textit{Generic} & -8.89 & 6.88 \\
  vanilla & 0.01 & \textit{Omniscient} & -8.67 & 6.87 \\
  vanilla & 0.01 & \textit{REVISE} & -8.65 & 6.8 \\
  full & 0.05 & \textit{ECCo} & -1.29 & 0.397 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.356}} \\
  full & 0.05 & \textit{Omniscient} & -5.08 & 5.09 \\
  full & 0.05 & \textit{REVISE} & $-5.91 \cdot 10^{27}$ & $1.36 \cdot 10^{28}$ \\
  vanilla & 0.05 & \textit{ECCo} & -9.35 & 7.32 \\
  vanilla & 0.05 & \textit{Generic} & -8.85 & 6.87 \\
  vanilla & 0.05 & \textit{Omniscient} & -8.7 & 6.96 \\
  vanilla & 0.05 & \textit{REVISE} & -8.52 & 6.76 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.2}} & \color{blue}{\textbf{0.383}} \\
  full & 0.1 & \textit{Generic} & -1.5 & 0.735 \\
  full & 0.1 & \textit{Omniscient} & -5.17 & 5.23 \\
  full & 0.1 & \textit{REVISE} & $-3.06 \cdot 10^{26}$ & $7.7 \cdot 10^{26}$ \\
  vanilla & 0.1 & \textit{ECCo} & -9.33 & 7.32 \\
  vanilla & 0.1 & \textit{Generic} & -8.88 & 6.86 \\
  vanilla & 0.1 & \textit{Omniscient} & -8.69 & 6.9 \\
  vanilla & 0.1 & \textit{REVISE} & -8.68 & 6.81 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.12}} & \color{blue}{\textbf{0.217}} \\
  full & 0.5 & \textit{Generic} & -1.21 & 0.352 \\
  full & 0.5 & \textit{Omniscient} & -5.09 & 5.12 \\
  full & 0.5 & \textit{REVISE} & $-5.97 \cdot 10^{27}$ & $1.37 \cdot 10^{28}$ \\
  vanilla & 0.5 & \textit{ECCo} & -9.35 & 7.3 \\
  vanilla & 0.5 & \textit{Generic} & -8.89 & 6.92 \\
  vanilla & 0.5 & \textit{Omniscient} & -8.68 & 6.93 \\
  vanilla & 0.5 & \textit{REVISE} & -8.53 & 6.75 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.1}} & \color{blue}{\textbf{0.163}} \\
  full & 1 & \textit{Generic} & -1.49 & 0.726 \\
  full & 1 & \textit{Omniscient} & -5.16 & 5.2 \\
  full & 1 & \textit{REVISE} & $-3.09 \cdot 10^{26}$ & $7.22 \cdot 10^{26}$ \\
  vanilla & 1 & \textit{ECCo} & -9.34 & 7.36 \\
  vanilla & 1 & \textit{Generic} & -8.86 & 6.85 \\
  vanilla & 1 & \textit{Omniscient} & -8.7 & 6.9 \\
  vanilla & 1 & \textit{REVISE} & -8.69 & 6.85 \\
  full & 5 & \textit{ECCo} & -1.75 & 0.154 \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.363}} \\
  full & 5 & \textit{Omniscient} & -5.14 & 5.16 \\
  full & 5 & \textit{REVISE} & $-1.1 \cdot 10^{28}$ & $2.5 \cdot 10^{28}$ \\
  vanilla & 5 & \textit{ECCo} & -9.36 & 7.32 \\
  vanilla & 5 & \textit{Generic} & -8.88 & 6.91 \\
  vanilla & 5 & \textit{Omniscient} & -8.7 & 6.93 \\
  vanilla & 5 & \textit{REVISE} & -8.52 & 6.73 \\
  full & 10 & \textit{ECCo} & $-1.02 \cdot 10^{6}$ & $2.32 \cdot 10^{6}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.49}} & \color{blue}{\textbf{0.702}} \\
  full & 10 & \textit{Omniscient} & -5.13 & 5.16 \\
  full & 10 & \textit{REVISE} & $-3.74 \cdot 10^{26}$ & $9.09 \cdot 10^{26}$ \\
  vanilla & 10 & \textit{ECCo} & -9.31 & 7.33 \\
  vanilla & 10 & \textit{Generic} & -8.87 & 6.86 \\
  vanilla & 10 & \textit{Omniscient} & -8.7 & 6.89 \\
  vanilla & 10 & \textit{REVISE} & -8.69 & 6.83 \\
  full & 15 & \textit{ECCo} & $-3.31 \cdot 10^{13}$ & $7.54 \cdot 10^{13}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.22}} & \color{blue}{\textbf{0.37}} \\
  full & 15 & \textit{Omniscient} & -5.2 & 5.23 \\
  full & 15 & \textit{REVISE} & $-9.01 \cdot 10^{27}$ & $2.06 \cdot 10^{28}$ \\
  vanilla & 15 & \textit{ECCo} & -9.38 & 7.34 \\
  vanilla & 15 & \textit{Generic} & -8.86 & 6.87 \\
  vanilla & 15 & \textit{Omniscient} & -8.69 & 6.96 \\
  vanilla & 15 & \textit{REVISE} & -8.51 & 6.73 \\\bottomrule

\end{longtable}

>>>>

\subsection{Encoding Actionability Constraints}\label{sec-constraints}

@@

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. Here it is
crucial that avoid penalizing implausibility that arises due to
mutability constraints. For any mutability-constrained feature \(d\)
this can be achieved by enforcing
\(\mathbf{x}^\prime[d] = \mathbf{x}[d]\) whenever the constraint is
violated, by adjusting either one to the other. Setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) intuitively corresponds to
imposing user requirements on the target domain; from a Bayesian
perspective it can be understood as assuming a point mass prior for
\(p(\mathbf{x})\) with respect to feature \(d\)\footnote{For additional
  detail, see Section~\ref{sec-app-constraints} of the supplementary
  appendix.}.

@@

\begin{itemize}
\tightlist
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\item
  \(\theta\): Model parameters (unspecified).
\item
  \(\Theta\): Matrix of parameters.
\end{itemize}

@@

\subsection{Details on Encoding Mutability
Constraints}\label{sec-app-constraints}

In Section~\ref{sec-constraints} we explained that to avoid penalizing
implausibility that arises due to mutability constraints for any feature
\(d\), we impose a point mass prior on \(p(\mathbf{x})\) for that
features. This is best explained through a simple example involving a
binary linear classifier.

Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\), with
\(y \in \{1,2\}\). Then for \(y^+ \sim p(y)=2\) we have:

\begin{equation}\phantomsection\label{eq-grad}{
\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \begin{pmatrix} 0 & 0 \\ \mathbf{x}^\prime[1] - \mathbf{x}[1] & \mathbf{x}^\prime[2] - \mathbf{x}[2] \end{pmatrix}
}\end{equation}

Contrastive divergence pushes down (up) on the coefficient
\(\Theta[y^+,d]\) by the amount that feature \(d\) that is lower
(higher) for counterfactuals than samples. This has the intended effect
of aligning counterfactual explanations with the data, as discussed
above.

Suppose, for example, that feature \(d=1\) represents \emph{age} and our
model predicts the likelihood of credit default:
\(y \in \{1:=\text{'no default'},2:=\text{'default'}\}\). Then if
\(\mathbf{x}^\prime[1] < \mathbf{x}[1]\), contrastive divergence will
push down on the corresponding coefficient,
\(\Theta[\text{'default'},\text{'age'}]\). Intuitively, this has the
effect of decreasing the effect of \emph{age} that the model attributes
to the log odds of default, i.e.~the model learns to higher default risk
younger individuals. But if \emph{age} is immutable, then this
corresponds to penalizing younger individuals for implausibility that
arises due to a feature that they can change about themselves. This is
precisely what we want to avoid, and therefore we impose no penalty at
all.

\subsection{Gradient}\label{gradient}

\subsection{Full Training Loop}\label{full-training-loop}

Next, we defined the training objective and the counterfactual
generator. Finally, we initialize the optimization state.

Finally, we visualize the results:

\section{Detailed Results}\label{detailed-results}

@@

@@

@@

\begin{table}

@@

\centering{

}

\end{table}%

\begin{table}

@@

\centering{

}

\end{table}%

@@

\centering{

}

\end{table}%

@@

\centering{

}

\end{table}%

### 2025-02-10 paper/sections/method.qmd
progress

## Encoding Actionability Constraints

@@

Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to align model representations with user requirements. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div. This is best explained through a simple example involving a linear classifier: $\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}$. Then for $y^+ \sim p(y)$ we have:

$$
\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
$$ {#eq-grad}

In other words, contrastive divergence will push down (up) on the coefficient $\theta[d]$ corresponding to a feature $d$ that is lower (higher) for counterfactuals than samples. This has the intended effect of aligning counterfactual explanations with the data, as discussed above.

Crucially though, we want to avoid penalizing implausibility that arises due to mutability constraints. Suppose, for example, that feature $d$ represents *age* and our model predicts the likelihood of credit default.

>>>>

## Encoding Actionability Constraints {#sec-constraints}

@@

Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to align model representations with user requirements. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div. Here it is crucial that avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^\prime[d] = \mathbf{x}[d]$ whenever the constraint is violated, by adjusting either one to the other. Setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ intuitively corresponds to imposing user requirements on the target domain; from a Bayesian perspective it can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$^[For additional detail, see @sec-app-constraints of the supplementary appendix.].
